\chapter{Executive Summary}

\paragraph{Background}

The performance of popular relational database systems reflects a failure to exploit the vast pool of optimisation techniques which are in essence extremely hard to include in a query compiler.

\emph{Voodoo} \cite{Pirk:2016:VVA:3007328.3007336} is an expressive vector algebra which when used as an intermediate representation for database queries has been shown to generate highly parallelisable OpenCL code, often outperforming existing state-of-the-art in-memory query processors.

\paragraph{Motivation}

The existing Voodoo implementation was only a back-end, not able to execute arbitrary queries or support arbitrary schemas. It consisted of several disjoint components which were highly coupled with specific benchmarking and OpenCL features. Furthermore, the code generation approach was fragile, had little testing or documentation, and did not allow for post generation optimisation.

\paragraph{Achievements}

By creating a front-end with Calcite, we allow any JDBC client to query data in a user-defined schema with common data types. For an arbitrary query, we can generate and optimise a logical plan and convert it to equivalent Voodoo algebra.

The new back-end makes novel use of the Clang AST to generate OpenCL code from Voodoo. This ensures syntactic correctness and allows for pre and post code generation analysis and optimisation.

Finally, we have significantly lowered the barrier to entry for future research on Voodoo by adopting an expandable design and providing precise testing and documentation which will allow a wider audience to contribute to it.

\paragraph{Applications}

High-performance in-memory databases have many applications in industry. Voodoo, while highly portable, has previously shown exceptional performance against popular systems which will have a significant impact on applications requiring large scale data processing.

%%%% START YEET
%                  ___           ___                   
%                 /\__\         /\__\                  
%      ___       /:/ _/_       /:/ _/_         ___     
%     /|  |     /:/ /\__\     /:/ /\__\       /\__\    
%    |:|  |    /:/ /:/ _/_   /:/ /:/ _/_     /:/  /    
%    |:|  |   /:/_/:/ /\__\ /:/_/:/ /\__\   /:/__/     
%  __|:|__|   \:\/:/ /:/  / \:\/:/ /:/  /  /::\  \     
% /::::\  \    \::/_/:/  /   \::/_/:/  /  /:/\:\  \    
% ~~~~\:\  \    \:\/:/  /     \:\/:/  /   \/__\:\  \   
%      \:\__\    \::/  /       \::/  /         \:\__\  
%       \/__/     \/__/         \/__/           \/__/  
%%%%  END YEET