\chapter{Introduction}

Use \url{https://docs.google.com/document/d/1tCYCOFDeL8E3USIOunRNrqh6ekP_-kAQVHMrNJ4jk04} for achievments. Try to back everything up in a quantifiable way...

\section{Motivation}

Very brief discussion of project history and background...

1. Not very useable because designed around TPC-H benchmarks, no front-end...

2. Code generation extremely difficult, based on strings...

3. Existing code incredibely dense and complex but sparsely documented...

\section{Objectives}

Main goal: make this usable for researchers: provide a front-end to what was just a benchmarking application, improve code generation...

1. Build a front-end to get SQL -> Logical plan -> Vooodoo for arbitrary schema, query.

2. Generate code using AST, to improve readability of generated code, and ease of implementation.

\section{Achievements}

1. Document and formalise installation...

1a. Contribute bugfixes required to get code to compile and run, update documentation

1b. Provide a dockerfile that prescribes the required setup

1c. Introduce CI that ensures the project can always build

2. Develop Calcite adapter that supports these SQL queries...

2a. Support arbitrary schemas with these types compared to before...

2b. Support arbitrary queries using these logical operators compared to before...

2c. Support rule-based logical optimisations using Calcite...

2d. Support enumerable operators using Calcite...

2e. Support integration with C++ printer implementation...

2f. Support integration with C++ AST implementation, by extending API...

2g. Provide a JDBC connection through Calcite

(2h. Provide a front-end that demonstrates the full process).

3. Develop a Clang-AST based implementation of Voodoo API.

3a. Support these Voodoo operators compared to before...

3b. Improve quality of implementation code as measured by this metric...

3c. Improve quality of generated code as measured by this metric...

3d. Improve ease-of-use as demonstrated by example (cross?), by providing AST components...

3e. Support future extension to C-like languages such as CUDA through generic processing class...
